---
title: "Brachy exploratory plots"
author: "Martin van Rongen"
date: "2019-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(GGally)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

### All traits

```{r, message = FALSE, warning = FALSE}
ril_phenotypes %>%
  select(nitrate_level, 
         z68_shoots, 
         flowering_time, 
         flowering_senescence_interval,
         lifespan, 
         z68_greenness, 
         senescence_height, 
         seed_weight_g) %>% 
  
  ggpairs(aes(colour = nitrate_level, alpha = 0.7))
```

Save to a PDF:

```{r, message = FALSE, warning = FALSE}
ggsave("ril_paired_comparisons.pdf", path = "../../figures/ril_phenotypes/", width = 20, height = 20)
```

### Traits LN vs HN

```{r, message = FALSE, warning = FALSE}
traits <- c("z68_shoots", 
            "flowering_time", 
            "flowering_senescence_interval", 
            "lifespan", 
            "z68_greenness", 
            "senescence_height", 
            "seed_weight_g")

for (i in traits){
  plot <- ril_phenotypes %>% 
    group_by(batch, nitrate_level, ril_id) %>% 
    summarise(avg_trait = mean(!!ensym(i))) %>% 
    spread(nitrate_level, avg_trait) %>% 
  
    ggplot(aes(x = LN, y = HN)) +
    labs(title = paste0("Correlation LN to HN for ", i)) +
    geom_point() +
    geom_smooth(method = "lm", colour = "black", size = 0.5) +
    geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
    geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
  
  print(plot)
}

```

