---
title: "Brachy exploratory plots"
author: "Martin van Rongen"
date: "2019-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(GGally)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

### All traits

```{r}
ril_phenotypes %>%
  select(nitrate_level, z68_shoots, flowering_time, flowering_senescence_interval, lifespan,z68_greenness, senescence_height, seed_weight_g) %>% 
  ggpairs(aes(colour = nitrate_level, alpha = 0.7))
```

Save to a PDF:

```{r}
ggsave("ril_paired_comparisons.pdf", path = "../../figures/ril_phenotypes/", width = 20, height = 20)
```

## Traits LN vs HN

### Branching

```{r}
ril_phenotypes %>% 
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots)) %>% 
  spread(nitrate_level, avg_z68_shoots) %>% 
  
  ggplot(aes(x = LN, y = HN)) +
  labs(title = "Mean branching LN vs HN") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = 0.5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```

### Flowering time

```{r}
ril_phenotypes %>% 
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(avg_flowering_time = mean(flowering_time)) %>% 
  spread(nitrate_level, avg_flowering_time) %>% 
  
  ggplot(aes(x = LN, y = HN)) +
  labs(title = "Mean flowering time LN vs HN") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = 0.5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```

### Flowering senescence interval

```{r}
ril_phenotypes %>% 
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(avg_flowering_senescence_interval = mean(flowering_senescence_interval)) %>% 
  spread(nitrate_level, avg_flowering_senescence_interval) %>% 
  
  ggplot(aes(x = LN, y = HN)) +
  labs(title = "Mean flowering senescence interval LN vs HN") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = 0.5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```
