---
title: "Brachy exploratory plots"
author: "Martin van Rongen"
date: "2019-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(GGally)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

```{r}
ril_phenotypes %>%
  select(nitrate_level, z68_shoots, flowering_time, flowering_senescence_interval, z68_greenness, senescence_height, seed_weight_g) %>% 
  ggpairs(aes(colour = nitrate_level, alpha = 0.7))
  

  ggsave("ril_paired_comparisons.pdf", path = "../../figures/ril_phenotypes/", width = 20, height = 20)
```

