---
title: "Branchy RIL branching traits"
author: "Martin van Rongen"
date: "2019-05-30"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

## HN/LN branching by batch

Scatter plot of number of shoots at `z68_shoots` by `ril_id` and `nitrate_level`:

```{r}
ril_phenotypes %>%
  group_by(batch, nitrate_level) %>% 
  filter(!is.na(z68_shoots)) %>% 

ggplot(., aes(x = ril_id, y = z68_shoots, colour = nitrate_level)) +
    labs(title = "Branch number at z68 stage", 
       subtitle = "Per batch by nitrate level", 
       x = "Genotype (by ril_id)", 
       y = "Branches") +
  geom_point() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.7) +
  facet_wrap(. ~ batch, ncol = 4) +
  
  theme(
    axis.text.x = element_blank()
  )
```

The plot above shows that overall the number of branches on LN is much lower than on HN. The variation of branch number appears to be lower on LN than on HN as well.

Checking this for each batch.

Furthermore, highlighting the parental and Bd21-3 filler lines. There seems to be a lot of variation in the level of branching within each batch for Bd21-3 (orange points).

This means that there is probably going to be a high level of variation in (mean) plasticity too.


```{r}
ril_phenotypes %>%
  filter(!is.na(z68_shoots)) %>% 
  mutate(ril_id = reorder(ril_id, z68_shoots)) %>% 

ggplot(., aes(x = ril_id, y = z68_shoots, colour = nitrate_level)) +
      labs(title = "Branch number at z68 stage", 
       subtitle = "Per batch split by nitrate level", 
       x = "Genotype (by ril_id)", 
       y = "Branches") +
  geom_point() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.7) +
  facet_grid(nitrate_level ~ batch) +
  
  theme(legend.position = "none",
        axis.text.x = element_blank())
```

### Mean branching per genotype on HN or LN

Calculating the average number of branches per genotype on HN and LN.

```{r}
plot_df <- ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  select(batch, id, ril_id, nitrate_level, z68_shoots) %>% 
  filter(!is.na(z68_shoots)) %>% 
  summarise(n = n(),
            avg_branching = mean(z68_shoots),
            sd = sd(z68_shoots),
            se = sd / sqrt(n),
            ci = 2 * se) %>% 
  do(
  plots = ggplot(., aes(x = fct_reorder(ril_id, avg_branching), y = avg_branching )) +
    labs(title = (paste0("Mean branching on ",.$nitrate_level)),
       subtitle = "Mean per genotype (with 95% CI)",
       x = "Genotype (reordered)", 
       y = "Branches (mean)") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_errorbar(aes(ymin = avg_branching - ci, ymax = avg_branching + ci), width = 0.1) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        legend.position = "none")
  )

plot_df$plots
```

## Branching plasticity

### Plasticity per genotype per batch

How about branching plasticity?
Plotting the plasticity (= number of branches on HN - number of branches on LN) and reordering by low to high plasticity. Using `geom_jitter()` since many RILs have the same plasticity (it's not a continuous variable).

The following lines are highlighted:

* Bd3-1 (magenta)
* Bd21 (blue)
* Bd21-3 (orange)

**Averaging multiple RIL entries**

> Each batch, nitrate_level, ril_id summarise with the mean. In most cases there's one data point. For ones that appear more than once (e.g. Bd21-3 or a filler replacement) there will be an average of those values. From *that* table you can spread by nitrate_level because in that table the ril_id will be unique per genotype per batch (CHECK if each ril_id occurs the same number of times as the number of batches).

> NOTE! This does not "save" the lost values from plants that died. For example, RIL025 in batch 1 on LN died, so an extra RIL025 was included in batch 3, but it's not possible to just substitute the value from batch 3 into batch 1 and then calculate the plasticity.


```{r results = "hide", fig.keep = "all"}

plot_df <- ril_phenotypes %>%
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(z68_shoots_mean = mean(z68_shoots)) %>% 
  spread(nitrate_level, z68_shoots_mean) %>% 
  mutate(plasticity = HN - LN) %>% 
  filter(!is.na(plasticity)) %>% 
  
  do(
plots = ggplot(., aes(x = reorder(ril_id, plasticity), y = plasticity)) + 
  labs(title = (paste0("Branching plasticity for batch ",.$batch)), 
       subtitle = "Reordered from low to high", 
       x = "Genotype (reordered)", 
       y = "Branching plasticity") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.7) +
  
  theme(axis.text.x = element_blank()))

plot_df$plots
```

**Comments**

The average branching plasticity for Bd21-3 varies somewhat between batches. Sometimes it is greater than Bd3-1, but sometimes it is less.

Branching plasticity in Bd3-1 is consistently higher than Bd21 across all tested batches (1-4).


### Overall plasticity per batch

To see if there's an overall batch effect on plasticity, the plasticity is plotted for each batch as a summary statistic in the form of a `boxplot`.

**Note:** the average number of shoots on HN/LN is calculated for RILs that occur more than once in each batch/treatment combination.

```{r}
ril_phenotypes %>%
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(z68_shoots_mean = mean(z68_shoots)) %>% 
  spread(nitrate_level, z68_shoots_mean) %>% 
  mutate(plasticity = HN - LN) %>% 
  filter(!is.na(plasticity)) %>% 
  
  ggplot(., aes(x = as.factor(batch), y = plasticity, fill = as.factor(batch))) +
    labs(title = "Branching plasticity by batch",
       x = "Batch number", 
       y = "Branching plasticity") +
  expand_limits(y = 0) +
  geom_boxplot() +
  
  theme(legend.position = "none")
```

**Comments**

The overall branching plasticity appears to be higher in batches 3 and 4, compared to batches 1 and 2.

### Mean plasticity across batches

```{r}
ril_phenotypes %>%
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(z68_shoots_mean = mean(z68_shoots)) %>% 
  spread(nitrate_level, z68_shoots_mean) %>% 
  mutate(plasticity = HN - LN) %>% 
  filter(!is.na(plasticity)) %>% 
  ungroup() %>% 
  group_by(ril_id) %>%
  summarise(n = n(),
            avg_plasticity = mean(plasticity),
            sd = sd(plasticity),
            se = sd / sqrt(n),
            ci = 2 * se) %>% 
  
  ggplot(., aes(x = fct_reorder(ril_id, avg_plasticity), y = avg_plasticity )) +
  labs(title = "Branching plasticity (low to high)",
       subtitle = "Mean per genotype (with 95% CI)",
       x = "Genotype (reordered)", 
       y = "Branching plasticity (mean)") +
  geom_jitter() +
  geom_errorbar(aes(ymin = avg_plasticity - ci, ymax = avg_plasticity + ci), width = 0.1) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

**Comments**

2019-04-30: There is considerable variation between batches, at least when only the first 4 batches are taken into account.

### Branching and position

```{r}
# by position e.g A1, H2
# separate columns, with leading zero
ril_phenotypes %>% 
  mutate(col_pos = str_remove(position_grid_location, 
                              "[A-Z]"),
         row_pos = str_remove(position_grid_location, 
                              "[0-9]+")) %>% 
  ggplot(aes(col_pos, row_pos, fill = z68_shoots)) +
    labs(title = "Branch number by position", 
       subtitle = "Per batch for each nitrate level", 
       x = "Column", 
       y = "Row") +
  geom_tile() +
  facet_grid(batch ~ nitrate_level)
```

**Comments**

Branching on HN is clearly higher than on LN, but again there does not seems to be an obvious pattern in relation to the grid. This suggests that there are no obvious positional effects to branching.

> Should follow up a bit more on this, maybe check the average across rows/columns to rule out edge effects.

