---
title: "Brachy RIL flowering"
author: "Martin van Rongen"
date: "2019-05-01"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r}
library(tidyverse)
library(corrr)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

## Flowering time

### Flowering time and branching per batch

Scatter plot of number of shoots at `z68_shoots` vs 'flowering_time' per `batch`. Using `geom_jitter` because otherwise many points overlap.

```{r}
ril_phenotypes %>%

ggplot(., aes(x = flowering_time, y = z68_shoots, colour = nitrate_level)) +
    labs(title = "Flowering time vs branching", 
       subtitle = "By batch and nitrate level", 
       x = "Flowering time (days)", 
       y = "Branches") +
  geom_jitter() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.3) +
  facet_wrap(. ~ batch, ncol = 4)
```

**Comments**

There seems to be a positive correlation between branching and flowering time on HN. Explore this further by looking at the average branching and flowering time.

### Flowering time and branching across batches

Using `geom_jitter` because otherwise many points overlap.

```{r}
ril_phenotypes %>%

ggplot(., aes(x = flowering_time, y = z68_shoots, colour = nitrate_level)) +
    labs(title = "Flowering time vs branching", 
       subtitle = "By nitrate level", 
       x = "Flowering time (days)", 
       y = "Branches") +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.5) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

### Average flowering time and branching

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_flowering_time = mean(flowering_time)) %>% 

ggplot(., aes(x = avg_flowering_time, y = avg_z68_shoots, colour = nitrate_level)) +
    labs(title = "Average flowering time vs average branching", 
       subtitle = "By nitrate level for each genotype", 
       x = "Flowering time (days)", 
       y = "Branches") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

### Average flowering time plasticity and branching plasticity

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_flowering_time = mean(flowering_time)) %>% 
  gather(trait, measurement, nitrate_level:avg_flowering_time) %>% 
  group_by(ril_id, trait) %>% 
  mutate(treatment = row_number()) %>% 
  mutate(treatment = ifelse(treatment == 1, "LN", "HN")) %>% 
  unite(trait_treatment, trait, treatment, sep = "_") %>% 
  spread(trait_treatment, measurement, fill = NA) %>%
  drop_na() %>% 
  mutate(branching_plasticity = as.numeric(avg_z68_shoots_HN) - as.numeric(avg_z68_shoots_LN),
         flowering_plasticity = as.numeric(avg_flowering_time_HN) - as.numeric(avg_flowering_time_LN)) %>%

ggplot(., aes(x = flowering_plasticity, y = branching_plasticity)) +
    labs(title = "Flowering time plasticity vs branching plasticity", 
       subtitle = "Average for each genotype", 
       x = "Flowering time plasticity", 
       y = "Branching plasticity") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  
  theme(
    legend.position = "none"
  )
```

### Average flowering time and SPAD across all batches

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(z68_greenness)) %>% 
  summarise(avg_greenness = mean(z68_greenness),
            avg_flowering_time = mean(flowering_time)) %>% 

ggplot(., aes(x = avg_flowering_time, y = avg_greenness, colour = nitrate_level)) +
    labs(title = "Average flowering time vs average SPAD", 
       subtitle = "By nitrate level for each genotype", 
       x = "Flowering time (days)", 
       y = "SPAD score") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

### Average flowering time and plant height across all batches

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(senescence_height)) %>% 
  summarise(avg_height = mean(senescence_height),
            avg_flowering_time = mean(flowering_time)) %>% 

ggplot(., aes(x = avg_flowering_time, y = avg_height, colour = nitrate_level)) +
    labs(title = "Average flowering time vs average height", 
       subtitle = "By nitrate level for each genotype", 
       x = "Flowering time (days)", 
       y = "Plant height (cm)") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

**Comments**

It is clear that Bd3-1 (in magenta) is much taller than Bd21 (in blue) on both LN and HN. There does not seem to be much of a correlation between plant height and flowering time on either `nitrate_level`.

### Flowering time plasticity

```{r}
ril_phenotypes %>%
  group_by(ril_id, batch, nitrate_level) %>% 
  summarise(avg_flowering_time = mean(flowering_time)) %>% 
  spread(nitrate_level, avg_flowering_time) %>% 
  mutate(flowering_time_plasticity = HN - LN) %>% 
  filter(!is.na(flowering_time_plasticity)) %>% 
  
  ggplot(., aes(x = ril_id, y = flowering_time_plasticity)) +
  labs(title = "Flowering time plasticity",
       subtitle = "Average for each genotype per batch",
       x = "Genotype (ril_id)", 
       y = "Flowering time plasticity (mean)") +
  geom_point() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

### Average flowering time plasticity

```{r}
ril_phenotypes %>%
  group_by(batch, nitrate_level, ril_id) %>% 
  summarise(avg_flowering_time = mean(flowering_time)) %>% 
  spread(nitrate_level, avg_flowering_time) %>% 
  mutate(flowering_time_plasticity = HN - LN) %>% 
  filter(!is.na(flowering_time_plasticity)) %>% 
  ungroup() %>% 
  group_by(ril_id,) %>%
  summarise(n = n(),
            avg_flowering_time_plasticity = mean(flowering_time_plasticity),
            sd = sd(flowering_time_plasticity),
            se = sd / sqrt(n),
            ci = 2 * se) %>% 
  
  ggplot(., aes(x = fct_reorder(ril_id, avg_flowering_time_plasticity), y = avg_flowering_time_plasticity )) +
  labs(title = "Flowering time plasticity (low to high)",
       subtitle = "Mean per genotype (with 95% CI)",
       x = "Genotype (reordered)", 
       y = "Flowering time plasticity (mean)") +
  geom_point() +
  geom_errorbar(aes(ymin = avg_flowering_time_plasticity - ci, ymax = avg_flowering_time_plasticity + ci), width = 0.1) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        legend.position = "none")
```

### Flowering time and position

It is possible that some traits are affected by the position in which the plants were grown.

To test this, traits can be plotted by position using `geom_tile`, which creates a sort of heat map of the trait based on position.

```{r}
# by position e.g A1, H2
# separate columns, with leading zero
ril_phenotypes %>% 
  mutate(col_pos = str_remove(position_grid_location, 
                              "[A-Z]"),
         row_pos = str_remove(position_grid_location, 
                              "[0-9]+")) %>% 
  ggplot(aes(col_pos, row_pos, fill = flowering_time)) +
    labs(title = "Flowering time by position", 
       subtitle = "Per batch for each nitrate level", 
       x = "Column", 
       y = "Row") +
  geom_tile() +
  facet_grid(batch ~ nitrate_level)
```

**Comments**

There does not seems to be any obvious pattern in the flowering time based on position. Overall flowering appears to occur faster on HN than it does on LN.


> Should follow up a bit more on this, maybe check the average across rows/columns to rule out edge effects.

