---
title: "Brachy RIL greenness traits"
author: "Martin van Rongen"
date: "2019-05-01"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

### Average flowering time and SPAD across all batches

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(z68_greenness)) %>% 
  summarise(avg_greenness = mean(z68_greenness),
            avg_flowering_time = mean(flowering_time)) %>% 

ggplot(., aes(x = avg_flowering_time, y = avg_greenness, colour = nitrate_level)) +
    labs(title = "Average flowering time vs average SPAD", 
       subtitle = "By nitrate level for each genotype", 
       x = "SPAD score", 
       y = "Flowering time (days)") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

## Greenness (SPAD)

### Per batch by nitrate level for each RIL

```{r}
ril_phenotypes %>% 
  filter(!is.na(z68_greenness)) %>% 
  
  ggplot(., aes(x = ril_id, y = z68_greenness, colour = nitrate_level)) +
    labs(title = "SPAD score per RIL",
         subtitle = "Per batch by nitrate level",
         x = "Genotype",
         y = "SPAD score") +
  geom_jitter() +
  facet_grid(batch ~ nitrate_level) +
  
  theme(legend.position = "none",
        axis.text.x = element_blank())
```

**Comments**

The SPAD scores appear to be pretty evenly distributed across the RILs.
They seem to be a little lower on LN than on HN.

Exploring this further.

### SPAD per batch by nitrate level

Checking the overall SPAD score per batch for each nitrate level.

```{r}
ril_phenotypes %>% 
  filter(!is.na(z68_greenness)) %>% 
  
  ggplot(., aes(x = nitrate_level, y = z68_greenness, fill = nitrate_level)) +
  labs(title = "SPAD score per batch",
         subtitle = "By nitrate level",
         x = "Nitrate level",
         y = "SPAD score") +
  geom_boxplot() +
  facet_grid(. ~ batch) +
  
  theme(legend.position = "none")
```

**Comments**

The SPAD score appears to be consistently lower on LN than on HN.
Across batches the overall SPAD score range is fairly consistent.

### SPAD score vs branching

Is there a correlation between SPAD score and number of branches?
Using `geom_jitter` because there are quite a few measurements that have the same SPAD score and these points then overlap.

```{r}
ril_phenotypes %>% 
  filter(!is.na(z68_greenness)) %>% 
  
  ggplot(., aes(x = z68_greenness, y = z68_shoots, colour = nitrate_level)) +
  labs(title = "Branch number vs SPAD score", 
       subtitle = "Per batch for each nitrate level", 
       x = "SPAD score", 
       y = "Branches") +
  geom_jitter() +
  facet_grid(batch ~ nitrate_level) +
  
  theme(legend.position = "none")
```

**Comments**

There does not seem to be a correlation between the number of branches a plant forms and its SPAD score.

### Mean SPAD score vs branches across batches

```{r}
ril_phenotypes %>%
  filter(!is.na(z68_shoots), !is.na(z68_greenness)) %>% 
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_z68_greenness = mean(z68_greenness)) %>% 
  
  ggplot(., aes(x = avg_z68_greenness, y = avg_z68_shoots, colour = nitrate_level)) +
  labs(title = "Branch number vs SPAD score", 
       subtitle = "Mean across batches for each nitrate level", 
       x = "Mean SPAD score", 
       y = "Branches (mean)") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  facet_grid(. ~ nitrate_level) +
  
  theme(legend.position = "none")
```

**Comments**

There definitely does not seem to be a correlation between branch number and SPAD score.
