---
title: "Brachy RIL poster"
author: "Martin van Rongen"
date: "2019-06-18"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(viridis)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

### Branching HN/LN
```{r}
ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  select(batch, id, ril_id, nitrate_level, z68_shoots) %>% 
  filter(!is.na(z68_shoots)) %>% 
  summarise(n = n(),
            avg_branching = mean(z68_shoots)) %>% 
  
  ggplot(., aes(x = fct_reorder(ril_id, avg_branching), y = avg_branching, colour = nitrate_level)) +
    labs(title = "Tiller number",
         x = "Genotype (reordered)", 
         y = "Tiller number") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_errorbar(aes(ymin = avg_branching, ymax = avg_branching), width = 0.1) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.2, 0.9))

ggsave("branching_HN_LN.pdf", width = 5, height = 5, useDingbats = FALSE)
```

### Branching plasticity
```{r}
ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots)) %>% 
  spread(nitrate_level, avg_z68_shoots) %>% 
  mutate(plasticity = HN - LN) %>% 
  filter(!is.na(plasticity)) %>% 
  ungroup() %>% 
  group_by(ril_id) %>%
  summarise(n = n(),
            avg_plasticity = mean(plasticity)) %>% 
  
  ggplot(., aes(x = fct_reorder(ril_id, avg_plasticity), y = avg_plasticity )) +
  labs(title = "Branching plasticity (HN - LN)",
       x = "Genotype (reordered)", 
       y = "Branching plasticity") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = c(0.2, 0.9))

ggsave("branching_plasticity.pdf", width = 5, height = 5, useDingbats = FALSE)
```

### Reaction-norm plots parental lines
```{r, message = FALSE, warning = FALSE}
ril_phenotypes %>% 
  filter(ril_id %in% c("Bd3-1", "Bd21")) %>% 
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(seed_weight_g)) %>% 
  summarise(n = n(),
            avg_trait = mean(seed_weight_g),
            sd = sd(seed_weight_g),
            se = sd / sqrt(n),
            ci = 2 * se) %>%
  
  ggplot(aes(x = nitrate_level, y = avg_trait, group = ril_id, colour = ril_id)) +
  labs(title = "Seed weight per plant (g)",
       x = "Nitrate level") +
  expand_limits(y = 0) +
  geom_point() +
  geom_errorbar(aes(ymin = avg_trait - ci, ymax = avg_trait + ci), width = 0.1) +
  geom_line() +
  facet_wrap(. ~ ril_id) +
  
    theme(legend.position = "none",
          axis.title.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.major.y = element_blank())

ggsave("response_seed_weight.pdf", width = 3, height = 3, useDingbats = FALSE)
```