---
title: "Brachy RIL life traits"
author: "Martin van Rongen"
date: "2019-05-01"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

## Time-related

### Life span

See how the RIL life span is distributed between treatments.

```{r}
ril_phenotypes %>% 
  
  ggplot(., aes(x = ril_id, y = lifespan, colour = nitrate_level)) +
  labs(title = "RIL life span", 
       subtitle = "Per genotype for each nitrate level", 
       x = "Genotype", 
       y = "Lifespan (days)") +
  geom_jitter() +
  facet_grid(. ~ nitrate_level) +
  
  theme(legend.position = "none",
        axis.text.x = element_blank())
```

**Comments**

The plants grown on LN appear to have a wider life span range than the ones grown on HN. Also, the life span of HN-grown plants appears to be higher.

Need to check if this trend is clear in all batches.

### Life span per batch

Checking the life span per batch

```{r}
ril_phenotypes %>% 
  filter(!is.na(lifespan)) %>% 
  
  ggplot(., aes(x = nitrate_level, y = lifespan, fill = nitrate_level)) +
  labs(title = "RIL life span", 
       subtitle = "Per batch for each nitrate level", 
       x = "Nitrate level", 
       y = "Lifespan (days)") +
  geom_boxplot() +
  facet_grid(. ~ batch) +
  
  theme(legend.position = "none")
```

**Comments**

It does look like the overall life span of plants grown on LN is shorter than that of plants grown on HN. This trend is visible in most batches, but not so in batch 1.

## Lifespan vs branching

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(z68_shoots) & !is.na(lifespan)) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
         avg_lifespan = mean(lifespan)) %>% 

ggplot(., aes(x = avg_lifespan, y = avg_z68_shoots, colour = nitrate_level)) +
    labs(title = "Branch number vs lifespan", 
       subtitle = "Mean per genotype by nitrate level", 
       x = "Lifespan (days)", 
       y = "Branches") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  #geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

## Life strategies

How much time is each genotype spending in each life phase?

```{r}
ril_phenotypes %>% 
  mutate(prop_sow_flower = flowering_time / lifespan,
         prop_flower_senescence = flowering_senescence_interval / lifespan) %>% 
  select(ril_id, nitrate_level, prop_sow_flower, prop_flower_senescence) %>% 
  mutate(ril_id = fct_reorder(ril_id, prop_sow_flower, .fun = mean)) %>% 
  gather(prop_sow_flower:prop_flower_senescence, key = "prop_trait", value = "proportion") %>% 
  group_by(ril_id, nitrate_level, prop_trait) %>% 
  summarise(avg_proportion = mean(proportion)) %>%
  mutate(avg_proportion_sow_flower = avg_proportion[prop_trait == "prop_sow_flower"]) %>%
  filter(nitrate_level == "HN") %>% 
  
  ggplot(aes(x = fct_reorder(ril_id, avg_proportion_sow_flower), y = avg_proportion, fill = prop_trait, colour = prop_trait)) +
    labs(title = "Time spent in each life stage", 
    subtitle = "Sowing to flowering and flowering to senescence", 
    x = "Genotype (reordered)", 
    y = "Proportion") +
  geom_bar(stat = "identity") +
  coord_flip() +
  
  theme(legend.position = "none",
        axis.text.y = element_blank())
```

### Flowering to senescence response to N

```{r}
ril_phenotypes %>% 
  group_by(nitrate_level, ril_id) %>% 
  summarise(avg_flowering_senescence_interval = mean(flowering_senescence_interval)) %>% 
  spread(nitrate_level, avg_flowering_senescence_interval) %>% 
  mutate(flowering_senescence_plasticity = HN - LN) %>% 
  
  ggplot(aes(x = fct_reorder(ril_id, flowering_senescence_plasticity), y = flowering_senescence_plasticity)) +
    labs(title = "Flowering-senescence interval plasticity", 
    subtitle = "By genotype", 
    x = "Genotype (reordered)", 
    y = "Flowering-senescence interval plasticity") +
  geom_point() +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  #geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  
  theme(axis.text.x = element_blank())
```

### Flowering time - senescence interval vs branching

```{r}
ril_phenotypes %>%
  group_by(ril_id, nitrate_level) %>% 
  filter(!is.na(z68_shoots) & !is.na(flowering_senescence_interval)) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
         avg_flowering_senescence_interval = mean(flowering_senescence_interval)) %>% 

ggplot(., aes(x = avg_flowering_senescence_interval, y = avg_z68_shoots, colour = nitrate_level)) +
    labs(title = "Branch number vs flowering-senescence interval", 
       subtitle = "Mean per genotype by nitrate level", 
       x = "Flowering-senescence interval (days)", 
       y = "Branches") +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  #geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", alpha = 0.7) +
  facet_wrap(. ~ nitrate_level) +
  
  theme(
    legend.position = "none"
  )
```

## Life stategies vs shoot branching plasticity

### Flowering time plasticity vs shoot branching

```{r}
ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_flowering_time = mean(flowering_time)) %>% 
  gather(avg_z68_shoots:avg_flowering_time, key = "avg_trait", value = "measurement") %>% 
  spread(nitrate_level, measurement) %>% 
  mutate(plasticity = HN - LN) %>% 
  select(-HN, -LN) %>% 
  spread(avg_trait, plasticity) %>% 
  rename_all(funs(str_replace(., "avg", "avg_plasticity"))) %>% 
  
  ggplot(aes(x = avg_plasticity_flowering_time, y = avg_plasticity_z68_shoots)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```


### Flowering-senescence interval vs shoot branching

```{r}
ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_flowering_senescence_interval = mean(flowering_senescence_interval)) %>% 
  gather(avg_z68_shoots:avg_flowering_senescence_interval, key = "avg_trait", value = "measurement") %>% 
  spread(nitrate_level, measurement) %>% 
  mutate(plasticity = HN - LN) %>% 
  select(-HN, -LN) %>% 
  spread(avg_trait, plasticity) %>% 
  rename_all(funs(str_replace(., "avg", "avg_plasticity"))) %>% 
  
  ggplot(aes(x = avg_plasticity_flowering_senescence_interval, y = avg_plasticity_z68_shoots)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```

### Life span plasticity vs shoot branching

```{r}
ril_phenotypes %>%
  group_by(nitrate_level, ril_id) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_lifespan = mean(lifespan)) %>% 
  gather(avg_z68_shoots:avg_lifespan, key = "avg_trait", value = "measurement") %>% 
  spread(nitrate_level, measurement) %>% 
  mutate(plasticity = HN - LN) %>% 
  select(-HN, -LN) %>% 
  spread(avg_trait, plasticity) %>% 
  rename_all(funs(str_replace(., "avg", "avg_plasticity"))) %>% 
  
  ggplot(aes(x = avg_plasticity_lifespan, y = avg_plasticity_z68_shoots)) +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7)
```

