---
title: "Brachy exploratory plots"
author: "Martin van Rongen"
date: "2019-05-01"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

**Note:** Exclude Bd21-3 filler plants?

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

## Seed weight vs branches

### Per batch by nitrate level

Checking if the total seed weight is correlated to the number of branches.

```{r}
ril_phenotypes %>%
  filter(!is.na(z68_shoots)) %>% 

ggplot(., aes(x = seed_weight_g, y = z68_shoots, colour = nitrate_level)) +
    labs(title = "Branch number vs seed weight",
         subtitle = "Per batch by nitrate level",
         x = "Seed weight (g)",
         y = "Branches") +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
  facet_grid(batch ~ nitrate_level) +
  
  theme(legend.position = "none")
```

**Comments**

There appears to be no correlation between seed weight per plant and the number of branches.

### Mean seed weight across batches

```{r}
ril_phenotypes %>%
  filter(!is.na(z68_shoots), !is.na(seed_weight_g)) %>% 
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_z68_shoots = mean(z68_shoots),
            avg_seed_weight_g = mean(seed_weight_g)) %>% 
  
  ggplot(., aes(x = avg_seed_weight_g, y = avg_z68_shoots, colour = nitrate_level)) +
  labs(title = "Branch number vs seed weight", 
       subtitle = "Mean across batches for each nitrate level", 
       x = "Mean seed weight (g)", 
       y = "Branches (mean)") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
    geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_grid(. ~ nitrate_level) +
  
  theme(legend.position = "none")
```

### Mean seed weight vs senescence

```{r}
ril_phenotypes %>%
  filter(!is.na(lifespan), !is.na(seed_weight_g)) %>% 
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_lifespan = mean(lifespan),
            avg_seed_weight_g = mean(seed_weight_g)) %>% 
  
  ggplot(., aes(x = avg_seed_weight_g, y = avg_lifespan, colour = nitrate_level)) +
  labs(title = "Seed weight vs lifespan", 
       subtitle = "Mean across batches for each nitrate level", 
       x = "Mean seed weight (g)", 
       y = "Lifespan (days)") +
  expand_limits(y = 0) +
  geom_jitter() +
  geom_smooth(method = "lm", colour = "black", size = .5) +
    geom_point(data = . %>%  filter(ril_id == "Bd3-1"), colour = "magenta", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21"), colour = "royalblue3", size = 3, alpha = 0.7) +
  geom_point(data = . %>%  filter(ril_id == "Bd21-3"), colour = "orange", size = 3, alpha = 0.7) +
  facet_grid(. ~ nitrate_level) +
  
  theme(legend.position = "none")
```

### Top-bottom yielders

```{r}
ril_seed_weight <- ril_phenotypes %>% 
  filter(!is.na(seed_weight_g)) %>% 
  group_by(ril_id, nitrate_level) %>% 
  summarise(avg_seed_weight = mean(seed_weight_g))
```

```{r}
most_seed_weight <- ril_seed_weight %>% 
  group_by(nitrate_level) %>% 
  top_n(10, avg_seed_weight) %>% 
  arrange(nitrate_level, desc(avg_seed_weight))

least_seed_weight <- ril_seed_weight %>% 
  group_by(nitrate_level) %>% 
  top_n(-10, avg_seed_weight) %>% 
  arrange(nitrate_level, desc(avg_seed_weight))

most_seed_weight
least_seed_weight
```


