---
title: "Brachy RIL data distribution"
author: "Martin van Rongen"
date: "2019-05-23"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created previously is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

## Traits

There are the following traits in the data set:

```{r}
names(ril_phenotypes)
```

## Histograms of data

### Branching

```{r}
ril_phenotypes %>%
  
  ggplot(., aes(x = z68_shoots, fill = nitrate_level)) +
      labs(title = "Branch number distribution", 
       subtitle = "Per batch by nitrate level", 
       x = "Branches", 
       y = "Count") +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  facet_wrap(. ~ batch)
```

### Flowering time

```{r}
ril_phenotypes %>%
  
  ggplot(., aes(x = flowering_time, fill = nitrate_level)) +
      labs(title = "Flowering time distribution", 
       subtitle = "Per batch by nitrate level", 
       x = "Flowering time (days)", 
       y = "Count") +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  facet_wrap(. ~ batch)
```

### Lifespan

```{r}
ril_phenotypes %>%
  
  ggplot(., aes(x = lifespan, fill = nitrate_level)) +
      labs(title = "Lifespan distribution", 
       subtitle = "Per batch by nitrate level", 
       x = "Life span (days)", 
       y = "Count") +
  geom_histogram(binwidth = 1, alpha = 0.7, position = "identity") +
  facet_wrap(. ~ batch)
```

### Plant height

```{r}
ril_phenotypes %>%
  
  ggplot(., aes(x = senescence_height, fill = nitrate_level)) +
      labs(title = "Plant height distribution", 
       subtitle = "Per batch by nitrate level", 
       x = "Plant height (cm)", 
       y = "Count") +
  geom_histogram(alpha = 0.7, position = "identity") +
  facet_wrap(. ~ batch)
```

### Seed weight

```{r}
ril_phenotypes %>%
  
  ggplot(., aes(x = seed_weight_g, fill = nitrate_level)) +
      labs(title = "Seed weight distribution", 
       subtitle = "Per batch by nitrate level", 
       x = "Seed weight (g)", 
       y = "Count") +
  geom_histogram(alpha = 0.7, position = "identity") +
  facet_wrap(. ~ batch)
```