---
title: "Brachy PCA analysis"
author: "Martin van Rongen"
date: "2019-05-03"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Loading libraries and settings

Load the relevant libraries and change some default plotting settings
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(visdat)
library(ggfortify)
library(broom)
theme_set(theme_bw())
scale_colour_discrete <- function(...) scale_color_brewer(palette = "Dark2")
scale_fill_discrete <- function(...) scale_fill_brewer(palette = "Dark2")
```

## Read in clean data

To ensure that the data is read in correctly, the `.csv` file that was created above is read in using the `read_csv` command. This way a tibble is created where the column specification is as basic as possible. This helps to avoid unnecessary factor conversions.

```{r}
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv")
```

Reorder the `nitrate_level`.

```{r}
ril_phenotypes <- ril_phenotypes %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))
```

How are the missing values distributed throughout the data set?

Note that removing the NAs basically eliminates batch 1.

```{r}
vis_miss(ril_phenotypes)
```

```{r}
pca_data <- ril_phenotypes %>%
  mutate(batch = as.factor(batch)) %>% 
  select(nitrate_level,
         batch,
         ril_id,
         z68_shoots,
         z68_greenness,
         flowering_time,
         flowering_senescence_interval,
         senescence_height,
         seed_weight_g,
         lifespan) %>%
  drop_na()
```

PCA analysis

* center = centering around the mean
* scale the data because the data are measured on different scales

> Colour the parents
> Connect the individuals with lines, do they cluster together?
> Store the PCA values as a trait in the original data set, you can perform QTLs & other analyses on them then.

```{r}
pca_data_num <- pca_data %>% 
  select(-(nitrate_level:ril_id)) %>% 
  prcomp(center = TRUE, scale. = TRUE)

autoplot(pca_data_num,
         data = pca_data,
         colour = "nitrate_level",
         loadings = TRUE,
         loadings.label = TRUE)
```

You get three things from your PCA:

1. PC scores
access from object
2. Eigenvalues (PC variance)

```{r}
str(pca_data_num)

# This gives the values of the PC
pca_data_num$x

# Bind the PC values to the original data set
bind_cols(pca_data, as_tibble(pca_data_num$x)) %>%
  # Create logical column to determine species for plotting purposes
  mutate(is_wt = ril_id == "Bd21-3") %>% 
  ggplot(aes(PC1, PC2, colour = is_wt)) +
  geom_point()

tidy(pca_data_num, matrix = "x")

# Eigenvalues = variance associated with each principal component
#pca_data_num$sdev
# Or in a tidy
tidy(pca_data_num, matrix = "pcs")

# Scree plot
tidy(pca_data_num, matrix = "pcs") %>% 
  ggplot(aes(PC, percent)) +
  geom_col() +
  geom_line(aes(y = cumulative))

# Return the weight/loading of the original variables on the new principal components.
tidy(pca_data_num, matrix = "variables")
```

### PCA plot `ggplot`

```{r}
# Create a matrix
pca_data <- ril_phenotypes %>%
  mutate(batch = as.factor(batch)) %>% 
  select(nitrate_level,
         batch,
         ril_id,
         z68_shoots,
         z68_greenness,
         flowering_time,
         flowering_senescence_interval,
         senescence_height,
         seed_weight_g,
         lifespan) %>%
  drop_na() %>% 
  group_by(nitrate_level, ril_id) %>% 
    summarise_at(vars(z68_shoots, z68_greenness, senescence_height, seed_weight_g, 
                    lifespan, flowering_senescence_interval, flowering_time),
               .funs = list(mean = ~ mean(., na.rm = TRUE))) %>% 
  unite("id", ril_id, nitrate_level, sep = "_")


pca_matrix <- pca_data %>% 
  select(-id) %>% 
  as.matrix()

rownames(pca_matrix) <- pca_data$id

# Perform the PCA
sample_pca <- prcomp(pca_matrix)

# Convert PCA matrix to tibble
pca_matrix <- pca_matrix %>% 
  as_tibble(rownames = "id") %>% 
  separate(id, c("ril_id", "nitrate_level"), sep = "_") %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))

# The PC scores are stored in the "x" value of the prcomp object, which is a matrix
pc_scores <- sample_pca$x %>%
  as_tibble(rownames = "id") %>% 
  separate(id, c("ril_id", "nitrate_level"), sep = "_") %>%
  mutate(nitrate_level = factor(nitrate_level, levels = c("LN", "HN")))

# Make PCA plot
pc_scores %>% 
  ggplot(aes(x = PC1, y = PC2, colour = nitrate_level)) +
  geom_point()

# PCA plot, connecting individual RILs
pc_scores %>% 
  ggplot(aes(x = PC1, y = PC2, colour = nitrate_level)) +
  geom_point() +
  geom_line(aes(group = interaction(ril_id)), colour = "black", alpha = 0.1)
```

### Trait loadings

```{r}
trait_loadings <- sample_pca$rotation %>% 
  as_tibble(rownames = "id")

ggplot(trait_loadings) +
  geom_segment(aes(x = 0, y = 0, xend = PC1, yend = PC2), 
               arrow = arrow(length = unit(0.1, "in")),
               colour = "brown") +
  geom_text(data = trait_loadings, 
            aes(x = PC1, y = PC2, label = id),
            nudge_y = 0.005, size = 3) +
  scale_x_continuous(expand = c(0.5, 0.03))
```



> Plot PC2 vs PC3 and see if you can find out what determines/separates them

> PCA, colour by nitrate, do the points separate

> How about by batch? Batch effects

