---
title: "Flowering"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r libraries, message = FALSE, warning = FALSE}
# load libraries
library(tidyverse)
library(patchwork)
library(broom)
library(lme4)

# change ggplot2 defaults
theme_set(theme_bw())
```

```{r functions}
#' Fit several relevant models to our data
#'
#' @param model_data a data.frame containing the data to be used in the model
#' @param trait a character vector containing the trait to be used as outcome variable
fit_lmms <- function(model_data, trait){
  
  if(!(trait %in% names(model_data))) stop("Trait does not exist")
  
  # define trait column
  model_data$trait <- model_data[[trait]]
  
  # check and remove missing values
  if(any(is.na(model_data$trait))){
    warning("There's missing values in the trait. They will be removed.\n")
    model_data <- model_data %>% drop_na(trait)
  }
  
  # fit and return models
  lmm_fits <- with(model_data,
                 list(
                   n_ril_int_batch = lmer(trait ~ nitrate_level  + (nitrate_level|ril_id) + (1|batch),
                                           control = lmerControl(optimizer ="Nelder_Mead")),
                   n_ril_int_batch_intn = lmer(trait ~ nitrate_level  + (nitrate_level|ril_id) + (nitrate_level|batch),
                                           control = lmerControl(optimizer ="Nelder_Mead")),
                   n_ril_batch_intn = lmer(trait ~ nitrate_level  + (1|ril_id) + (nitrate_level|batch),
                                           control = lmerControl(optimizer ="Nelder_Mead")),
                   n_ril_int = lmer(trait ~ nitrate_level  + (nitrate_level|ril_id)),
                   n_ril_batch = lmer(trait ~ nitrate_level  + (1|ril_id) + (1|batch)),
                   n_ril = lmer(trait ~ nitrate_level  + (1|ril_id)),
                   n_batch = lmer(trait ~ nitrate_level + (1|batch)),
                   ril = lmer(trait ~ (1|ril_id)),
                   n = lm(trait ~ nitrate_level),
                   lm_fit = lm(trait ~ nitrate_level*ril_id)
                 ))
  return(lmm_fits)
}
```


```{r}
# read original data - column types are explicitly defined
all_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv",
                           col_types = cols(
                             .default = col_double(),
                             sheet = col_character(),
                             identifier = col_character(),
                             block = col_character(),
                             shelf = col_character(),
                             nitrate_level = col_factor(levels = c("LN", "HN")),
                             id = col_character(),
                             batch = col_factor(),
                             position_grid_location = col_character(),
                             ril_id = col_character(),
                             sow_date = col_date(format = "%Y-%m-%d"),
                             z68_date = col_date(format = "%Y-%m-%d"),
                             senescence_date = col_date(format = "%Y-%m-%d")
                           )) %>% 
  # exclude filler line
  filter(ril_id != "Bd21-3") %>% 
  # retain only lines that have data in both nitrate treatments
  group_by(ril_id) %>% 
  filter(n_distinct(nitrate_level) == 2) %>% 
  ungroup()

# Split data into:
# test - will not be touched on until the end
# training - will be used to play around with different models
test <- all_phenotypes %>% filter(batch %in% c("1", "2"))
training <- all_phenotypes %>% filter(!(batch %in% c("1", "2")))
```

Prepare the data for the model

```{r}
# Clean data for model
model_data <- training %>% 
  # select variables of interest - rename trait of interest
  select(batch, nitrate_level, ril_id, trait_raw = flowering_time) %>% 
  # remove missing values
  drop_na() %>% 
  # scale it
  mutate(trait = (trait_raw - mean(trait_raw))/sd(trait_raw))

# quick check for outliers (using residuals from linear model):
lm(trait ~ nitrate_level:ril_id, data = model_data) %>% plot(which = 1)
```

No obvious outliers.

```{r}
model_data %>% filter(abs(trait) >= 5)
```


```{r}
# Create table that will be used to get model predictions
model_pred <- model_data %>% 
  distinct(ril_id, nitrate_level)

# fit models
model_fits <- fit_lmms(model_data, "trait")
```


## Explore model outputs

Looking at the GxE model:

```{r}
summary(model_fits$n_ril_int)
```

We get warning of _boundary fit_, which we can see is because there's basically no GxE as far as the model is concerned.

```{r}
model_pred %>% 
  mutate(pred = predict(model_fits$n_ril_int, model_pred),
         mean = predict(model_fits$lm_fit, model_pred)) %>% 
  gather("model", "prediction", pred, mean) %>% 
  mutate(x = factor(interaction(model, nitrate_level), 
                    levels = c("mean.LN", "pred.LN", "pred.HN", "mean.HN"))) %>% 
  ggplot(aes(x, prediction)) +
  geom_line(aes(group = ril_id), alpha = 0.5)
```

Basically the model predicts no GxE, and this results in quite a bit of shrinkage on both N treatments. 

It's a good idea to see what the plasticity distribution is in the population:

```{r}
model_pred %>% 
  mutate(mean = predict(model_fits$lm_fit, model_pred)) %>% 
  spread(nitrate_level, mean) %>% 
  mutate(plas = HN - LN) %>% 
  ggplot(aes(plas)) + stat_ecdf()
```

It's positive, with inter-quartile range between ~0.5 and 1 SDs. 
I assume that there's too much within-RIL variance, which makes the model not trust that this variation is meaningful. 
Long lower tail, suggesting some lines flowered _later_ on LN (which seems strange...).

Let's look at AIC across models:

```{r}
model_fits %>% 
  map_dbl(AIC) %>% 
  sort() %>% 
  enframe(name = "model", value = "AIC") %>% 
  mutate(dAIC = AIC - min(AIC),
         relative_likelihood = exp(-dAIC/2),
         AIC_weights = relative_likelihood/sum(relative_likelihood))
```

Saturated model is at the top, but not very different from the other models. 

Models including _batch_ seem to come on top there, we should look at this variation:

```{r}
model_data %>% 
  group_by(ril_id, batch, nitrate_level) %>% 
  summarise(trait = mean(trait)) %>% 
  ggplot(aes(interaction(batch, nitrate_level), trait)) + 
  geom_line(aes(group = ril_id), alpha = 0.2) +
  geom_point(stat = "summary", fun.y = "mean", aes(colour = nitrate_level), size = 3)
```

Yes, substantial and consistent batch effect, with batches 3-4 earlier on both nitrate treatments.


## Conclusion

- There is substantial batch effect; could it be micro-environmental cues?
- There seems to be no significant GxE for the trait, as the model with `nitrate|ril_id` has correlation close to 1. The model seemed to converge, although we should be careful with inferences there. 

```{r}
summary(model_fits$n_ril_int_batch_intn)
```

Percentage of variance:

```{r}
model_fits$n_ril_int_batch_intn %>% 
  VarCorr() %>% 
  as_tibble() %>% filter(is.na(var2)) %>% 
  mutate(pct_var = vcov/sum(vcov)*100)
```

We can look at the residual and predictive-observed plots:

```{r}
# residual plot
p1 <- model_fits$n_ril_int_batch_intn %>% 
  augment(model_data) %>% 
  ggplot(aes(.fitted, .resid)) +
  geom_point(aes(colour = nitrate_level)) + 
  stat_smooth(se = FALSE) +
  geom_hline(yintercept = 0, linetype = 2, size = 1) +
  labs(title = "Residual plot")

p2 <- model_fits$n_ril_int_batch_intn %>% 
  augment(model_data) %>% 
  ggplot(aes(.fitted, trait)) +
  geom_point(aes(colour = nitrate_level), show.legend = FALSE) + 
  geom_abline(linetype = 2, size = 1) +
  labs(title = "Predicted vs. Observed")

p1 + p2 + plot_layout(ncol = 1)
```

We can also look at the prediction on the test set:

```{r}
# add predictions to test set
test <- test %>% 
  mutate(predicted = predict(model_fits$n_ril_int_batch_intn, newdata = test, allow.new.levels = TRUE),
         observed = (flowering_time - mean(model_data$trait_raw)) / sd(model_data$trait_raw)) %>% 
  drop_na(observed)

# predicted versus observed
test %>% 
  ggplot(aes(predicted, observed)) +
  geom_point(aes(colour = nitrate_level)) +
  geom_abline() +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = 0, linetype = 2) +
  theme_classic()

# difference between predicted and observed
test %>% 
  ggplot(aes(predicted - observed)) +
  geom_density(aes(colour = nitrate_level)) +
  geom_vline(xintercept = 0, linetype = 2, size = 1)

test %>% 
  group_by(nitrate_level) %>% 
  summarise(spearman = cor(predicted, observed, method = "spearman"),
            pearson = cor(predicted, observed, method = "pearson"))
```


