---
title: "Mixed model QTL scans"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
# load libraries
library(qtl2)
library(qtl2helper) # remotes::install_github("tavareshugo/qtl2helper")
library(tidyverse)
library(patchwork)
library(lme4)

# change ggplot2 defaults
theme_set(theme_bw() + theme(text = element_text(size = 12)))
```


## Summary

Trying to run QTL scans using the non-summarised dataset, i.e. with all individual data. 


## Read data

```{r}
# read data as a cross2 object
ril_cross <- read_cross2("../../data/external/Cui2012_qtl2cross.zip")

# insert pseudo-markers for evenly spaced scan
geno_map <- insert_pseudomarkers(ril_cross$gmap, step = 0.5)

# calculate genotype probabilities - assume 1% error rate
geno_prob <- calc_genoprob(ril_cross, geno_map, error_prob = 0.01, cores = 3)

# read trait data
# read original data - column types are explicitly defined
ril_phenotypes <- read_csv("../../data/processed/ril_phenotypes/ril_phenotypic_data.csv",
                           col_types = cols(
                             .default = col_double(),
                             sheet = col_character(),
                             identifier = col_character(),
                             block = col_character(),
                             shelf = col_character(),
                             nitrate_level = col_factor(levels = c("LN", "HN")),
                             id = col_character(),
                             batch = col_factor(),
                             position_grid_location = col_character(),
                             ril_id = col_character(),
                             sow_date = col_date(format = "%Y-%m-%d"),
                             z68_date = col_date(format = "%Y-%m-%d"),
                             senescence_date = col_date(format = "%Y-%m-%d")
                           )) %>% 
  # retain only lines with genotype
  filter(ril_id %in% ind_ids(ril_cross)) %>% 
  # retain only lines that have data in both nitrate treatments
  group_by(ril_id) %>% 
  filter(n_distinct(nitrate_level) == 2) %>% 
  ungroup()
```


## Shoot branching

- We're removing batches 1 & 2 from this analysis, later we can try including batch effects, I guess.
- We've also concluded from previous models that there's no LN variation, so we fit the model only to HN data.

```{r}
# prepare data for the model
model_data <- ril_phenotypes %>% 
  # exclude batches 1 & 2
  filter(!(batch %in% c("1", "2")) & nitrate_level == "HN") %>% 
  # remove missing values
  drop_na(ril_id, z68_shoots) %>% 
  # create a generic name for the trait of focus
  rename(trait = z68_shoots)

# Variables for mixed model
PHEN <- model_data$trait
ID <- model_data$ril_id
```

First we fit the null model. From previous exploration we include a term for batch and do not include interaction.

```{r}
# Fit null model
fit_null <- lme4::lmer(PHEN ~ (1|ID))
summary(fit_null)
```

```{r}
# get genotypes for all lines (including replicates)
impute_genos <- viterbi(ril_cross, geno_map, error_prob = 0.01, cores = 3)

all_gen <- impute_genos[model_data$ril_id, ]

# fit genetic models to each marker
lod_scores <- suppressMessages(
  lapply(
    all_gen, function(chr){ 
      lods <- lapply(
        dimnames(chr)[[2]], function(marker){
          
          # Get genotype matrix
          GEN <- chr[, marker]
          
          # convert it to binary values (dummy variable)
          GEN <- (GEN == 2)*1
          
          tryCatch(
            {
              # Fit each of the models
              # Full model including interaction term
              #fit_full <- lme4::lmer(PHEN ~ COV + GEN + GEN:COV + (1|ID) + (1|BATCH))
              
              # Genetic model including genetic but no interaction term
              fit_gen <- lme4::lmer(PHEN ~ GEN + (1|ID))
              
              # Output a table with LOD scores for each
              data.frame(
                # LOD_full = (logLik(fit_full) - logLik(fit_null))/log(10),
                # LOD_GxE = (logLik(fit_full) - logLik(fit_gen))/log(10),
                LOD_G = (logLik(fit_gen) - logLik(fit_null))/log(10)
              )
              
            }, error=function(err) data.frame(LOD_full = NA, LOD_GxE = NA, LOD_G = NA))    
        }
      )
      
      names(lods) <- dimnames(chr)[[2]]
      
      return(lods)
    }
  )
)

# Convert list to tibble
lod_scores <- lod_scores %>% 
  unlist(recursive = FALSE) %>% 
  dplyr::bind_rows(.id = "marker") %>% 
  mutate(marker = str_remove(marker, "^[1-5]\\."))

# Get markers
markers <- lapply(geno_map, 
                  function(i){
                    data.frame(marker = names(i), bp = as.numeric(i), stringsAsFactors = FALSE)
                  }) %>% 
  dplyr::bind_rows(.id = "chrom")

# Join the marker and test tables
lod_scores <- merge(markers, lod_scores, by = "marker", all = TRUE)
```

Visualise:

```{r}
lod_scores %>% 
  ggplot(aes(bp, LOD_G)) +
  geom_line() +
  facet_grid(. ~ chrom, scales = "free_x", space = "free_x") +
  scale_x_continuous(breaks = seq(0, 500, 100)) +
  labs(x = "Position (cM)")
```


## Comparing QTL model strategies

- mean trait values per RIL
- median trait values per RIL
- BLUP trait values per RIL
- the mixed model above

Let's get the traits summarised:

```{r}
ril_summary <- model_data %>% 
  group_by(ril_id) %>% 
  summarise(mean_trait = mean(trait),
            median_trait = median(trait))

# add the BLUP estimates
ril_summary$blup_trait <- lmer(trait ~ (1|ril_id), data = model_data) %>% 
  predict(newdata = ril_summary)
```

We can plot them together:

```{r}
ril_summary %>% 
  mutate(ril_id = reorder(ril_id, mean_trait)) %>% 
  gather("metric", "value", -ril_id) %>% 
  ggplot(aes(ril_id, value, colour = metric)) +
  geom_point() +
  theme(axis.text.x = element_blank())
```

And run the scan with the the three different summary metrics:

```{r}
temp <- add_pheno(ril_cross, ril_summary)

scan1(geno_prob, temp$pheno) %>% tidy(map = geno_map) %>% 
  ggplot(aes(pos, LOD)) +
  geom_line() +
  facet_grid(pheno ~ chrom, scales = "free_x", space = "free_x")
```



## Flowering time

- We're removing batches 1 & 2 from this analysis, later we can try including batch effects, I guess.
- We're modelling HN and LN together, with no GxE component, since preliminary analysis suggests there isn't any.

```{r}
# prepare data for the model
model_data <- ril_phenotypes %>% 
  # exclude batches 1 & 2
  filter(!(batch %in% c("1", "2"))) %>% 
  # remove missing values
  drop_na(ril_id, nitrate_level, flowering_time) %>% 
  # create a generic name for the trait of focus
  rename(trait = flowering_time)

# Variables for mixed model
PHEN <- model_data$trait
COV <- factor(model_data$nitrate_level, levels = c("LN", "HN"))
BATCH <- factor(model_data$batch)
ID <- model_data$ril_id
```

First we fit the null model. From previous exploration we include a term for batch and do not include interaction.

```{r}
# Fit null model
fit_null <- lme4::lmer(PHEN ~ COV + (1|ID) + (1|BATCH))
summary(fit_null)
```

Now fitting the genetic model:

```{r}
# get genotypes for all lines (including replicates)
impute_genos <- viterbi(ril_cross, geno_map, error_prob = 0.01, cores = 3)

all_gen <- impute_genos[model_data$ril_id, ]

# fit genetic models to each marker
lod_scores <- suppressMessages(
  lapply(
    all_gen, function(chr){ 
      lods <- lapply(
        dimnames(chr)[[2]], function(marker){
          
          # Get genotype matrix
          GEN <- chr[, marker]
          
          # convert it to binary values (dummy variable)
          GEN <- (GEN == 2)*1
          
          tryCatch(
            {
              # Fit each of the models
              # Full model including interaction term
              #fit_full <- lme4::lmer(PHEN ~ COV + GEN + GEN:COV + (1|ID) + (1|BATCH))
              
              # Genetic model including genetic but no interaction term
              fit_gen <- lme4::lmer(PHEN ~ COV + GEN + (1|ID) + (1|BATCH))
              
              # Output a table with LOD scores for each
              data.frame(
                # LOD_full = (logLik(fit_full) - logLik(fit_null))/log(10),
                # LOD_GxE = (logLik(fit_full) - logLik(fit_gen))/log(10),
                LOD_G = (logLik(fit_gen) - logLik(fit_null))/log(10)
              )
              
            }, error=function(err) data.frame(LOD_full = NA, LOD_GxE = NA, LOD_G = NA))    
        }
      )
      
      names(lods) <- dimnames(chr)[[2]]
      
      return(lods)
    }
  )
)

# Convert list to tibble
lod_scores <- lod_scores %>% 
  unlist(recursive = FALSE) %>% 
  dplyr::bind_rows(.id = "marker") %>% 
  mutate(marker = str_remove(marker, "^[1-5]\\."))

# Get markers
markers <- lapply(geno_map, 
                  function(i){
                    data.frame(marker = names(i), bp = as.numeric(i), stringsAsFactors = FALSE)
                  }) %>% 
  dplyr::bind_rows(.id = "chrom")

# Join the marker and test tables
lod_scores <- merge(markers, lod_scores, by = "marker", all = TRUE)
```


Visualise:

```{r}
lod_scores %>% 
  ggplot(aes(bp, LOD_G)) +
  geom_line() +
  facet_grid(. ~ chrom, scales = "free_x", space = "free_x") +
  scale_x_continuous(breaks = seq(0, 500, 100)) +
  labs(x = "Position (cM)")
```
